Aircraft Wildlife Collisions

### 2021-07-20 draft_final.Rmd
Added plot description and interpretation as well as summary.

```{r echo=FALSE, warning=FALSE,message = FALSE,fig.align = "center", fig.asp=0.5}

@@

set.seed(43)
top50 <- machines[order(-machines$n)[1:50],] %>% 
  mutate(angle = 30 * sample(-2:2, n(), replace = TRUE, prob = c(1, 1, 4, 1, 1)))
top50_aircrafts_wordcloud <- ggplot(top50) +
  aes(label = atype, size=n, color=ac_mass, angle=angle) +
  geom_text_wordcloud_area(
    #mask = png::readPNG(source="flugzeug6-3.png"),
    #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
    rm_outside = TRUE) +
  scale_size_area(max_size = 3) +
  scale_color_brewer(palette="BuPu") +#scale_color_continuous("Blues") +
  #ggtitle("top 50 aircraft models causing strikes")
  #theme(plot.title = element_text(vjust=-20)) +
  theme.title+
  labs(
    title="Top 50 aircrafts involved in collisions",
    subtitle = "Model types by number of collisions"
       )

@@

The left plot shows that especially the B-737 class (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727. "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas. The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes. Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this. But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.

From the right plot we can see, that aircraft models with higher max speed are also able to fly higher. This can be also seen from the linear model, that has been estimated by geom_smooth. The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high. In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height. We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent. The group with ac_mass 27001-272000 kg) is the most prominent group under the Top50 aircrafts. Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.

The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all aircraft models (n=`r nrow(machines)`. Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.

@@

From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines. We can also see that lighter machines tend to have less engines first group includes nearly only machines with ac_mass=1, second group is mixed with ac_mass 2-4 and the last two groups include mostly aircrafts with ac_mass=4. Looking at the contingency table also confirms that:

```{r echo=FALSE, fig.asp=0.5}
# create some fake data (2x2, since we're building a confusion matrix) 
dat <- matrix(table(machines$num_engs, machines$ac_mass), nrow = 4, dimnames=list(c(1:4), c(1:5)))
# now set the names *of the dimensions* (not the row/colnames)
names(dimnames(dat)) <- c("num_engs", "ac_mass")
dat
```
Interesting is that for aircrafts with num_engs=1, there is no model with a speed > 200 and height > 10000. So we could conclude from that this is the maximum height and speed possible with one engine.

@@

# 3. Analysis of influences for birds_struck, birds_seen and effect
In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect). We will first start with the analysis of strikes.

@@

### Birds struck out of birds seen
birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
  scale_fill_brewer(palette="Blues") +
  theme(axis.text.y=element_blank(), 
        axis.title.y=element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90),
        legend.position="left")+
  theme.title+
  labs(title = "Birds struck overview",
       subtitle = "bird struck out of birds seen")

@@

The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100". So **most often we do not see a birds before a strike** happens and most often 1 bird is struck. But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones. If the pilot sees birds, **quite often less birds were hit than seen**.  Interesting here is, that if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen. The event of hitting less birds than seen will later be analyzed more in detail.

The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens during Landing, Climb and Take-Off phase.

From the bottom right plot we can see that strikes happen most often during the Day with over 10500 occurences followed by Night with about 4500 occurences. Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).

@@

birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
  aes(x=birds_struck, fill=sky)+
  scale_fill_manual(values = color_pal_sky)+
  geom_bar(position = "fill")+
  geom_text(aes(label=ifelse(round(..count../struck_sky_sums,2) > 0.1,round(..count../struck_sky_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 90)) +
  theme.title+
  labs(title = "Birds struck comparisons",
       subtitle = "Cloudiness by birds struck")

@@

birds_strike_ac_mass <- ggplot(birds_struck_minorities)+
aes(x=ac_mass, fill=birds_struck)+
  scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
  geom_bar(position = "fill") +
  geom_text(aes(label=ifelse(round(..count../ac_struck_sums[1:14],2) > 0.1,round(..count../ac_struck_sums[1:14],2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  theme.title+
  labs(subtitle = "Birds struck (subset) by aircraft mass")

@@

From the plot on the left we can see that with **increasing number of birds_struck the proportion of overcast increases** and the proportion of no clouds decreases. The proportion of some clouds is more or less the same across all groups. From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy.

TODO: add Chi-squared independence test for sky and birds struck
http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r

We will now have a look on the height variable together with phase of flight.

### 3.1.1 Closer look on height of collisions

The top plot shows a boxplot of height for flight phases, which are above ground and on the bottom we will have a closer look on the height for the flight phases "En Route" and "Descent", because they show the most variability in height. We defined Climb, En Route, Descent, Approach as phases which are above ground level.

@@

#plot height density for Descent and EnRoute
density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  scale_fill_brewer(palette='Pastel2') +
  geom_density(adjust=1.5,alpha=.8) +
  xlim(0,20000)

@@

The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
The other two boxplots for "En Route" and "Descent" are more far away from ground. Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.

We are performing a t.test to compare the two groups and check for equality.

@@

The T-Test also show that the group means are different from each other. But surprisingly the **mean in group "Descent" is higher** than mean of "En Route". One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach or during flight in general. So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route". The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent". So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**. And especially a negative angle increases the probability of hitting birds, based on the findings here and from 3.1.1 by comparing the number of collisions by phase of flight.

### 3.1.2 density birdsstuck by height for flight phases above ground

@@

```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
#could include only birds from vys categorization
#plot height density for Descent and EnRoute
density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
  aes(x = height, group=birds_struck,fill=birds_struck)+
  geom_density(adjust=1.5,alpha=.6) +
  scale_fill_brewer(palette="Blues") +
  theme(legend.position = "right") + 
  theme.title+
  xlim(0, 3500) +
  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")
density
```

The density plot shows a **high peek at the beginning at about 100 foot** for all categories of birds_struck greater 0. This is, because most of the collisions happen near ground level at flight phases Take-Off, Climb, Approach and Landing Roll (see 3.1 Number of collisions by pflight phase). For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 150 foot to 500 foot. Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**. For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here. Unfortunately we have mostly missing values in the variable species for these observations, so we can not get any information about the birds species.

@@

there are only two levels in the data available for birds seen and we added one level for the missing values. The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot). At the bottom you can see two spineplots that show the two continuous variables height and speed by the number of birds seen.

```{r echo=FALSE, fig.asp=0.4}
#relationship of amount of birds seen with clouds?

time_sums <- rep(addmargins(table(birds$birds_seen, birds$time_of_day),1)[4,],each=3)
birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
  aes(x=time_of_day, fill=birds_seen)+
  geom_bar(position = "fill") +
  geom_text(aes(label=ifelse(round(..count../time_sums,2) > 0.1,round(..count../time_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  scale_fill_brewer(palette="BuGn") +
  theme(legend.position="none", axis.title.y = element_blank()) +
  theme.title+
  labs(title="Variables influencing birds seen",
       subtitle = "Proportion of birds seen by time")

@@

Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is 0.217+0.069=0.286. The contingency table looks like the following:

```{r echo=FALSE, fig.asp=0.5}
round(addmargins(prop.table(table(birds$birds_seen, birds$time_of_day),2),2,FUN = mean),3) 
```

The proportion of overall **birds seen at night is much lower** compared to the other times (Dawn, Day, Dusk). And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).

Sky: The proportion of **birds seen increases with cloudiness**. This is a **contradiction** to what we would have expected. We would have thought that with increasing cloudiness the amount of birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds. So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky. Another reason could be that birds in groups fly more often when its cloudy.

The higher the speed & height, the less likely to see a bird before collision: 
* for height > 2000 foot, the probability to see a birds before collision decreases 
* for speed > 140 knots, the probability to see a bird before collision decreases

## 3.3 Influences when birdsstruck < birds seen

In the following we will have a closer look on a self-generated variable, which compares the amount of birds seen with the amount of birds_struck. It will be true if the number of birds struck is smaller than birds seen and false if not. In the beginning we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck. By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.

@@

less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,],each=2)

# phase of flight by birdsstruck < birds seen
seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt),]) +
  aes(x=phase_of_flt, fill=less_strike_than_seen) +
  scale_fill_manual(values = color_pal_TF) +
  geom_bar(position = "fill") +
 # geom_text(aes(label=..count..),
  geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) > 0.1,round(..count../less_strike_sums[2:16],2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  coord_flip() +
  theme(legend.position="right", axis.title.y = element_blank())+
  theme.title+
  labs(title="Variables influencing less strikes than seen",
       subtitle = "Proportion of less strike than seen \nby phase of flight",
       fill='less strike \nthan seen'
       )

@@

Comparing less strike than seen across phases of flight we can see, that "En Route" it is about 12% more likely to strike less birds than seen compared to "Descent". So it is not even less likely to strike birds in general, but also to strike less birds than seen.
The num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife. This might be due to better maneuvarability of the lighter aircrafts. Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines.

@@

The top plot shows the different proportions of effect for each category of birds_struck. The Bottom left plot shows the proportions of each effect by the aircraft mass and on the bottom right we compare effect with birds_seen or not seen.

@@

birds_struck: The greater the number of birds struck, the more likely that the collision has an effect on the aircraft. Especially the proportion of "Precautionary Landing" increases. For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off".

ac_mass: the higher the aircraft mass the less likely the collision has an effect on the aircraft.

birds_seenYes: The "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and Precautionary Landing and Aborted Take-off are a little higher. We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike. But from the description of the dataset we can not really find out what is meant by the effect variable. Does "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision. When birds were seen we would expect to have less amount of effect on the aircraft, cause he could try to dodge the birds or abort take off/ land precautionary.

@@

>>>>

```{r echo=FALSE, warning=FALSE, message = FALSE,fig.align = "center", fig.asp=0.5}

@@

set.seed(43)
top50 <- machines[order(-machines$n)[1:50],] %>% 
  mutate(angle = 30 * sample(-2:2, n(), replace = TRUE, prob = c(1, 1, 4, 1, 1)))
top50_aircrafts_wordcloud <- ggplot(top50) +
  aes(label = atype, size=n, color=ac_mass)+#, angle=angle) +
  geom_text_wordcloud_area(
    #mask = png::readPNG(source="flugzeug6-3.png"),
    #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
    rm_outside = TRUE) +
  scale_size_area(max_size = 3) +
  scale_color_brewer(palette="BuPu") +
  #ggtitle("top 50 aircraft models causing strikes")
  #theme(plot.title = element_text(vjust=-20)) +
  theme.title+
  labs(
    title="Top 50 aircrafts involved in collisions",
    subtitle = "Model types by number of collisions"
       )

@@

The left plot shows that especially the **B-737 class** (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727. "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas. The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes. The B-737 is a narrow-body airliner produced by Boeing at its Renton Factory in Washington and was developed for short and thin routes. Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this. But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.

From the right plot we can see, that aircraft **models with higher max speed are also able to fly higher**. This can be also seen from the linear model, that has been estimated by geom_smooth. The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high. In addition we can see via the coloring by ac_mass that **lighter aircrafts tend to have less max_speed and max_height**. We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the Top50. The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the Top50 aircrafts. Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.

The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all `r nrow(machines)` aircraft models. Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.

@@

From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines. We can also see that lighter machines tend to have less engines: 
* first group includes nearly only machines with aircraft mass = 1 (less than 2250 kg), 
* second group is mixed with aircraft mass 2-4 (2251-272000 kg), 
* the last two groups include mostly aircrafts with mass = 5 (27001-272000 kg).

Looking at the contingency table also confirms that:

```{r echo=FALSE, fig.asp=0.5, fig.align="center"}
# create some fake data (2x2, since we're building a confusion matrix) 
dat <- matrix(table(machines$num_engs, machines$ac_mass), nrow = 4, dimnames=list(c(1:4), c(1:5)))
# now set the names *of the dimensions* (not the row/colnames)
names(dimnames(dat)) <- c("num_engs", "ac_mass")
dat
```
Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot. So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.

@@

# 3. Analysis of influences for birds_struck, birds_seen and effect
In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect). Most of the time we will use barplots with position fill, mosaicplot and spineplots as we want to compare proportions and check for dependencies. We will first start with the analysis of strikes.

@@

### Birds struck out of birds seen
birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
  geom_mosaic(aes(x = product(birds_struck,birds_seen), fill=birds_struck)) +
  scale_fill_brewer(palette="Blues") +
  #scale_fill_manual(values=brewer.pal(5,"Blues"))+
  theme(
    axis.text.y=element_blank(), 
        axis.title.y=element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90),
        legend.position="left")+
  theme.title+
  labs(title = "Birds struck overview",
       subtitle = "bird struck out of birds seen")

@@

The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100". So **most often we do not see birds before a strike** happens and most often 1 bird is struck. But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones. If the pilot sees birds, **quite often less birds were hit than seen**. For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen. The event of hitting less birds than seen will later be analyzed more in detail in chapter 3.3.

The top right plot shows that collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll. The amount of collision in Approach phase is highest and about 2 times higher than the other start and end phases. During the middle of the flight in phases "En Route" and "Descent" less collisions occur.

From the bottom right plot we can see that strikes happen most often during the Day with over 10500 collisions followed by Night with about 4500. Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).

@@

birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
  aes(x=birds_struck, fill=sky)+
  scale_fill_manual(values = color_pal_sky)+
  geom_bar(position = "fill")+
  geom_text(aes(label=ifelse(round(..count../struck_sky_sums,2) > 0.1,round(..count../struck_sky_sums,2) , "")),
            stat='count',
            size = 2.3,
            position=position_fill(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 90)) +
  theme.title+
  labs(title = "Birds struck comparisons",
       subtitle = "Cloudiness by birds struck")

@@

birds_strike_ac_mass <- ggplot(birds_struck_minorities)+
aes(x=ac_mass, fill=birds_struck)+
  #scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
  scale_fill_manual(values=brewer.pal(5,"Blues")[c(1,4,5)]) +
  geom_bar(position = "fill") +
  geom_text(aes(label=ifelse(round(..count../ac_struck_sums[1:14],2) > 0.1,round(..count../ac_struck_sums[1:14],2) , "")),
            stat='count',
            size = 2.3,
            position=position_fill(vjust = 0.5))+
  theme.title+
  labs(subtitle = "Birds struck (subset) by aircraft mass")

@@

From the plot on the left we can see that with **increasing number of birds_struck the proportion of overcast increases** and the proportion of no clouds decreases. The proportion of some clouds is more or less the same across all groups. From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy or the pilot as well as the birds are not able to see each other because of the clouds and so nobody can try to prevent the collision by dodging the other one.

On the right side we can see a subset of the birds_struck groups, which focuses on the minor categories in our dataset. From that we can see that the aircraft mass has an influence on the amount of birds struck. For the lightest aircraft class up to 2250 kg the proportion of 0 birds struck is very high, which means that the pilot has seen a bird and was likely to hit one, but he did not. For this category we would expect that lighter machines are able to initiate evasive maneuvers as they are more agile. The relationship of striking less birds than seen will be analyzed more in detail in chapter 3.3. 
It also stands out, that the proportion of striking over 100 birds is significantly higher for the heaviest aircrafts above 272000 kg. This will most likely be due to the fact that heavier machines are also larger in size and thus they will hit more birds when flying through a flock of birds compared to lighter and thus smaller machines.

### 3.1.1 Height of collisions above ground

We will now have a look on the height variable together with phase of flight. The top plot shows a boxplot of height for flight phases, which are above ground. We defined Climb, En Route, Descent, Approach as phases which are above ground level. On the bottom plot we will then have a closer look on the height for the flight phases "En Route" and "Descent" by visualizing the density.

@@

#plot height density for Descent and EnRoute
density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  scale_fill_manual(values=brewer.pal(4,"Pastel2")[2:3]) +
  geom_density(adjust=1.5,alpha=.8) +
  xlim(0,20000)

@@

The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data. Thus, we will focus these two. Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile. We will investigate more the differences of the two distributions by using a density plot.

Both flight phases have the highest density for lower heights up to about 5000 foot. The curve for En Route is much steeper, but a little more narrow compared to "Descent". Both density cross at about 3750 which is the highest point of "Descent" phase. From that height upwards the probability to strike a birds is higher when being in Descent phase. This will be most likely because of the angle of the aircraft at which it is flying. In phase En Route it will be most likely parallel to the ground, so the height of the airplane does not change during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for Approach phase.

We are performing a t.test to compare the two group height ("Descent" and "En Route"), so that we can confirm that both distributions are unequal.

@@

The T-Test also shows that the group means are different from each other and that the **mean in group "Descent" is higher** than mean of "En Route". One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase, where the aircraft is loosing height to prepare for approach or during flight in general. So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route". The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent". So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**. And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 by comparing the number of collisions by phase of flight.

### 3.1.2 Density of height by birds_struck for flight phases above ground

@@

```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
#could include only birds from vys categorization
#plot height density for Descent and EnRoute
density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
  aes(x = height, group=birds_struck,fill=birds_struck)+
  geom_density(adjust=1.5,alpha=.6) +
  scale_fill_brewer(palette="Blues") +
  theme(legend.position = "right") + 
  theme.title+
  xlim(0, 3500) +
  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")
density

#above_ground[above_ground$height>1500 & above_ground$height < 2000 & above_ground$birds_struck=="Over 100",]
```

The density plot shows that the probability for striking a bird is highes near ground level. This plot shows a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0. So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases Take-Off, Climb, Approach and Landing Roll have the highest number of strikes (see 3.1 Number of collisions by pflight phase). For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 foot to 500 foot. Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**. For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here. So for this category we have the about the same probability of hitting birds near ground level and at about 1800 foot. One of the involved species here was a **flock of about 500 gulls**. Unfortunately the other observations had missing values in the variable "species", so we can not get more information about the species involved.

@@

There originally were only two levels in the data available for birds seen and we added one level for the missing values. The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot). At the bottom you can see two spineplots that show the two continuous variables height and speed by the number of birds seen.

```{r echo=FALSE, fig.asp=0.4}
#relationship of amount of birds seen with clouds?
time_sums <- rep(addmargins(table(birds$birds_seen, birds$time_of_day),1)[4,],each=3)
birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
  aes(x=time_of_day, fill=birds_seen)+
  geom_bar(position = "fill") +
  geom_text(aes(label=ifelse(round(..count../time_sums,2) > 0.1,round(..count../time_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  scale_fill_brewer(palette="BuGn") +
  theme(legend.position="none", axis.title.y = element_blank()) +
  theme.title+
  labs(title="Variables influencing birds seen",
       subtitle = "Proportion of birds seen by time")

@@

Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the average proportion of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285. So most often pilots do not see birds before a collision. The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day, Dusk).

Sky: The proportion of **wildlife seen increases with cloudiness**. This is a **contradiction** to what we would have expected. We would have thought that with increasing cloudiness the amount of wildlife/birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds. So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky. Another reason could be that birds fly more often when its cloudy.

From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the speed & height, its becoming even less likely to see birds before a collision. For height greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.

## 3.3 Influences when number of birds struck is less than birds seen

In the following we will have a closer look on a self-generated variable, which compares the amount of birds seen with the amount of birds_struck. It will be true if the number of birds struck is smaller than birds seen and false if not. In the beginning (chapter 3.1.1) we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck. By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.

@@

less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,3:8],each=2)

# phase of flight by birdsstruck < birds seen
seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt) & as.integer(seen_vs_strike$phase_of_flt)>2,]) +
  aes(x=phase_of_flt, fill=less_strike_than_seen) +
  scale_fill_manual(values = color_pal_TF) +
  geom_bar(position = "fill") +
 # geom_text(aes(label=..count..),
 # geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) >
  geom_text(aes(label=ifelse(round(..count../less_strike_sums,2) >
  #0.1,round(..count../less_strike_sums[2:16],2) , "")),
                               0.1,round(..count../less_strike_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  coord_flip() +
  theme(legend.position="right", axis.title.y = element_blank())+
  theme.title+
  labs(title="Variables influencing less strikes than seen",
       subtitle = "Proportion of less strike than seen \nby phase of flight",
       fill='less strike \nthan seen'
       )

@@

For the left barplot we observe a similar relationship compared to the plot of number of collisions by flight phases from chapter 3.1.1, over the start, middle and landing of the flight. The proportion of less strikes is increasing from the start of the flight (Take-Off Run) and reaches the highest proportion of 0.69 in the phase "En Route" and then decreases again, where it reaches a minimum of 0.48 in the last phase "Landing Roll". So we see again, that the angle of the flight must have an influence on striking less birds than seen.
The number of engines and aircraft mass also have an influence of how much birds are hit out of the birds seen.
Lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife. This might be again due to better maneuverability of the lighter and thus smaller aircrafts. Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines which matches also with the findings of the unique aircraft models from chapter 2.

@@

The description of the effect variable from the dataset is insifficient. We can not really find out what is meant by the effect variable. Does for example "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision. We interpreted this variable as it really is the effect on the aircraft because a collision has occurred.
The top plot shows the different proportions of effect for each category of birds_struck. The Bottom left plot shows the proportions of each effect by the aircraft mass and on the bottom right we compare effect with birds_seen or not seen.

@@

The greater the number of birds struck, the more likely that the collision has an effect on the aircraft. Especially the proportion of "Precautionary Landing" increases. For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off", which makes sense, because the pilot had to intervene somehow to prevent the collision. For the aircraft mass, we can see that the larger the aircraft mass the less likely the collision has an effect on the aircraft. This might be due to the fact, that heavier machines are more robust. But for aircraft with an mass=5 (greater 272000 kg) we can see a an increase in the proportion of effect, which might be due to the fact that these machines have a higher probability to strike over 100 birds, which will then also have an effect on the most robust aircrafts.

Having a look on the last plot, we can see that the "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and "Engine Shutdown" and "Other" are a little higher. We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike, because the pilot could then try to prevent the collision.

@@

B-737 class causes the most strikes. It includes aircrafts from Boeing which were developed for short and thin routes. Aircraft models with higher max speed are also able to fly higher. Lighter machines tend to have less engines. With one engine max. speed and max. height is limited to 200 knots and 10000 foot

Most of the time we do not see wildlife/birds before a collision and if the pilot sees something, he most often hits less amount of birds/wildlife than seen. Collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll. 
The proportion of overcast increases with the number of birds struck, this might be because of birds flying in flocks more often when it is cloudy or due to the lack of vision which prevents evasive maneuvers.
The mass of the aircraft has an influence on the amount of birds struck, lighter machines are most likely more agile and smaller in size, thus they can better initiate evasive maneuvers, especially compared to machines with a mass greater 275000 kg, which have an about three times higher proportion of striking over 100 birds.
Bird strikes have a higher probability to occur near ground level and the probability of hitting birds is dependent on the phase of the flight. The probability to hit a bird is higher when the aircraft is having an negative angle and looses height.

Most often we do not see birds before a collision and at night its even less likely to see birds. Surprisingly the proportion of wildlife seen increases with cloudiness, this phenomenon needs to be further analyzed. Greater speed and height decreases the probability to see a bird.

If the aircraft is flying parallel to the ground the proportion of pilots striking less birds out of the seen is higher. The greater the number of engines and aircraft mass the less likely less birds are hit out of the birds seen.

The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
The larger the aircraft mass the less likely the collision has an effect on the aircraft.
There is less proportion of effect when birds were seen, which is a contradiction to our intrepretation of the effect variable.
